Nonparametric Block Bootstrap Kolmogorov-Smirnov Goodness-of-Fit Test

### 2024-05-08 manuscript/manuscript.tex
application and conclusion section

The Kolmogorov-Smirnov (KS) test is a widely used statistical test that
assesses the conformity of a sample to a specified distribution. Its efficacy,
however, diminishes with serially dependent data and when parameters
within the hypothesized distribution are unknown. For independent data,
parametric and nonparametric bootstrap procedure are available to adjust for
estimated parameters. For serially dependent stationary data, parametric
bootstrap has been developed with a working serial dependence structure. A
counterpart for the nonparametric bootstrap approach, which needs a bias
correction, has not been studied. Addressing this gap, our study introduces a
bias correction method employing a nonparametric block bootstrap, which
approximates the distribution of the KS statistic accounting for unspecified
serial dependence and unspecified parameters. We assess its effectiveness
through simulations, scrutinizing both its size and power. The practicality of
our method is further illustrated with an examination of stock returns from the 
S\&P 500 index, showcasing its utility in real-world
applications.

@@

The standard one-sample Kolmogorov-Smirnov (KS) test is widely
recognized as an effective good-of-fit test for continuous distributions.
Consider $X_1, ..., X_n$, a random sample of size~$n$ from some continuous
distribution and the null hypothesis $H_0$ that $X_i$'s follow a specific
hypothesized distribution~$F(x; \theta_0)$, where $\theta_0$ are the 
specified parameters
of the hypothesized distribution family.
If we let $F_n(t) = \sum_{i=1}^n I(X_i \le t) / n$ be the empirical cumulative
distribution function of the sample, where $I(\cdot)$ is the indicator
function, the KS test statistic takes the form
\begin{equation}
  \label{eq:ks_standard}
  T_n = \sqrt{n} \sup_x | F_{n}(x) - F(x; \theta_0)|.
\end{equation}
As sample size $n\to \infty$, the distribution of $T_n$ converges to that of the
absolute value of standard Brownian bridge, which is known as the Kolmogorov
distribution \citep{stephens1974edf}. This distribution function is
precisely computable using contemporary statistical software
\citep{marsaglia2003evaluating}. The versatility of the KS test allows its
application across various domains, such as analyzing cosmic microwave
background radiation \citep{naess2012application}, monitoring the count rate of
radioactive data \citep{aslam2020introducing}, gear condition monitoring
\citep{andrade2001gear}, studying images of breast cancer tumors
\citep{demidenko2004kolmogorov}, and examining financial markets
\citep{lux2001turbulence}.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width = \textwidth]{figures/normal}
  \includegraphics[width = \textwidth]{figures/gamma}
  \caption{Q-Q plots of the p-values testing that a time series
    have marginal normal distribution with true data generating distribution
    being $N(8,8)$ (upper) and marginal gamma distribution with true data
    generating distribution $\Gamma(8,1)$ (lower)}
  \label{fig:normal}
\end{figure}

@@

Using the \textsl{tseries} package \citep{tseries}, 
we gathered daily closing prices for the S\&P 500 index from January 1st, 2020
to December 31st, 2023. Daily returns were obtained by taking the difference in
the logarithm of the closing prices. The Kendall's $\tau$ for the series
was about -0.0298, indicating a very slight serial dependence.
Using our method, we tested that the daily returns are normally
distributed and Student's $t$ distributed 
with degrees of freedom $v \in \{30, 20, 10, 5, 4, 3, 2, 1\}$.
We tested the same hypotheses using 1) the
semiparametric bootstrap method shown by \citet{zeimbekakis2022misuses} which 
accounts for both serial dependence and unspecified parameters, 2)
nonparametric basic bootstrap bias correction developed by 
\citet{babu2004goodness}, and 3) the parametric bootstrap method shown
by \citet{zeimbekakis2022misuses}.
We used $B = 10,000$ bootstrap 
replicates in this analysis.
The p-values for these tests are highlighted in 
Table~\ref{table:SP5004}.

\input{tables/sp500_pval_4y}

While differences in p-values among different methods are expected, our analysis
reveals numerous instances where conclusions at the 0.05 significance level
using our method differ from those obtained using the methods proposed by
\citet{babu2004goodness} and \citet{zeimbekakis2022misuses}.
For example, when assessing the S\&P 500 series, our method fails to reject the
hypothesis that the series follows a Student's $t$ distribution with degrees of
freedom $v = 5$ with a p-value of 0.061. In contrast, 
the semiparametric method yields a p-value of 0.0192,
the basic nonparametric method
yields a p-value of 0.0222, and the parametric method yields a p-value of
0.0141. Although such differences are not guaranteed, this
instance highlights a disagreement in conclusions among methods for the S\&P 500
index.

\section{Concluding Remarks}
\label{sec:conclusion}

The Kolmogorov-Smirnov test stands as a crucial tool for researchers across
diverse fields, enabling comparisons of population distributions. This paper
introduces a variant of the test tailored for serially dependent data without
necessitating parameter specification.
Using simulation, we have shown that given a large enough sample of a time 
series and if the serial dependence is not too extreme, the block 
bootstrap KS test can appropriately fail to reject the null hypothesis when the
series follows the null distribution. In addition, when the marginal
distribution is not
the one hypothesized, the test is powerful. Notably, it demands a larger sample
size compared to basic bootstrap methods to avoid both type I and type II 
errors. Unsurprisingly,
the method performs better as the temporal dependence gets weaker. Through
simulation studies with Normal and Gamma-distributed data, we have shown that
that the method is robust.

We have also 
demonstrated possible applications of this method to financial data.
Through comparison with the nonparametric
bootstrap bias correction developed by \citet{babu2004goodness}
and the parametric and semiparametric
bootstrap methods developed by \citet{zeimbekakis2022misuses}, we have 
underscored the importance of accounting for serial
dependence in choosing an appropriate methodology.
This method holds relevance for studies aiming to assess if a serially dependent
time series adheres to a hypothesized distribution in scenarios where parameters
are unknown or unspecified. Future studies could investigate
the method's performance on different marginal distributions such
as the generalized extreme value distribution and
different dependence structures such as MA, ARMA, and ARIMA
processes. Additionally, one could propose a similar method
for comparing two serially dependent methods. One could also apply the method
showcased in this paper to fields like earthquake prediction and astronomy.

>>>>

The Kolmogorov--Smirnov (KS) test is a widely used statistical test that
assesses the conformity of a sample to a specified distribution. Its efficacy,
however, diminishes with serially dependent data and when parameters
within the hypothesized distribution are unknown. For independent data,
parametric and nonparametric bootstrap procedure are available to adjust for
estimated parameters. For serially dependent stationary data, parametric
bootstrap has been developed with a working serial dependence structure. A
counterpart for the nonparametric bootstrap approach, which needs a bias
correction, has not been studied. Addressing this gap, our study introduces a
bias correction method employing a nonparametric block bootstrap, which
approximates the distribution of the KS statistic accounting for unspecified
serial dependence and unspecified parameters. We assess its effectiveness
through simulations, scrutinizing both its size and power. The practicality of
our method is further illustrated with an examination of stock returns from the 
S\&P 500 index, showcasing its utility in real-world
applications.

@@

The standard one-sample Kolmogorov--Smirnov (KS) test is widely
recognized as an effective good-of-fit test for continuous distributions.
Consider $X_1, ..., X_n$, a random sample of size~$n$ from some continuous
distribution and the null hypothesis $H_0$ that $X_i$'s follow a specific
hypothesized distribution~$F(x; \theta_0)$, where $\theta_0$ are the 
specified parameters
of the hypothesized distribution family.
If we let $F_n(t) = \sum_{i=1}^n I(X_i \le t) / n$ be the empirical cumulative
distribution function of the sample, where $I(\cdot)$ is the indicator
function, the KS test statistic takes the form
\begin{equation}
  \label{eq:ks_standard}
  T_n = \sqrt{n} \sup_x | F_{n}(x) - F(x; \theta_0)|.
\end{equation}
As sample size $n\to \infty$, the distribution of $T_n$ converges to that of the
absolute value of standard Brownian bridge, which is known as the Kolmogorov
distribution \citep{stephens1974edf}. This distribution function is
precisely computable using contemporary statistical software
\citep{marsaglia2003evaluating}. The versatility of the KS test allows its
application across various domains, such as analyzing cosmic microwave
background radiation \citep{naess2012application}, monitoring the count rate of
radioactive data \citep{aslam2020introducing}, gear condition monitoring
\citep{andrade2001gear}, studying images of breast cancer tumors
\citep{demidenko2004kolmogorov}, and examining financial markets
\citep{lux2001turbulence}.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width = .9\textwidth]{figures/normal}
  \includegraphics[width = .9\textwidth]{figures/gamma}
  \caption{Q-Q plots of the p-values testing that a time series
    have marginal normal distribution with true data generating distribution
    being $N(8,8)$ (upper) and marginal gamma distribution with true data
    generating distribution $\Gamma(8,1)$ (lower)}
  \label{fig:qq}
\end{figure}

@@

Using the \textsl{tseries} package \citep{tseries}, 
we gathered daily closing prices for the S\&P 500 index from January 1st, 2020
to December 31st, 2023. Daily returns were obtained by taking the difference in
the logarithm of the closing prices. 
It is known as stylized facts that daily return series often exhibts nonlinear
dependence and unconditional heavy tails \citep[e.g.,][]{ryden1998stylized,
  cont2001empirical}. The Kendall's $\tau$ for the series is
was about $-0.0298$, indicating a fairly weak serial linear correlation, but
certainly not meaning absence of serial dependence.
Using our method, we tested that the daily returns are normally
distributed and Student's $t$ distributed 
with degrees of freedom $\nu \in \{30, 20, 10, 5, 4, 3, 2, 1\}$
with the proposed nonparametric block bootstrap (NPBB) method.
For comparison, we tested the same hypotheses using 1) the
semiparametric bootstrap (SPB) method of \citet{zeimbekakis2022misuses},
which accounts for both serial dependence and unspecified parameters;
2) nonparametric bootstrap (NPB) bias correction method of
\citet{babu2004goodness}; and and 3) the parametric bootstrap (PB) method shown
by \citet{zeimbekakis2022misuses}. The latter of the two methods discard the
serial dependence. In all methods, we used $B = 10,000$ bootstrap replicates.
The p-values for these tests are summarized in 
Table~\ref{table:SP5004}.

\begin{table}[ht]
\centering
\caption{P-values for 4 years of S\&P 500 stock return 
  data using different durations
  and different degrees of freedom for Student's $t$ distribution.
  NPBB: nonparametric block bootstrap;
  SPB: semiparametric bootstrap;
  NPB: nonparametric bootstrap;
  PB: parametric bootstrap.} 
\label{table:SP5004}
\begin{tabular}{ccccc}
  \toprule
$\nu$ & NPBB & SPB & NPB & PB\\ 
  \midrule
$\infty$ & 0.0001 & 0.0000 & 0.0000 & 0.0000 \\ 
  30 & 0.0019 & 0.0000 & 0.0000 & 0.0000 \\ 
  20 & 0.0027 & 0.0000 & 0.0000 & 0.0000 \\ 
  10 & 0.0125 & 0.0002 & 0.0003 & 0.0003 \\ 
  5 & 0.0610 & 0.0192 & 0.0222 & 0.0141 \\ 
  4 & 0.1037 & 0.0733 & 0.0574 & 0.0499 \\ 
  3 & 0.3133 & 0.3000 & 0.2759 & 0.2686 \\ 
  2 & 0.0418 & 0.0461 & 0.0377 & 0.0443 \\ 
  1 & 0.0000 & 0.0000 & 0.0000 & 0.0000 \\ 
  \bottomrule
\end{tabular}
\end{table}

While differences in p-values among different methods are expected, our analysis
reveals some instances where conclusions at the 0.05 significance level
using our method differ from those obtained using the methods proposed by
\citet{babu2004goodness} and \citet{zeimbekakis2022misuses}.
For example, our NPBB method fails to reject the
hypothesis that the series follows a Student's $t$ distribution with degrees of
freedom $v = 5$ with a p-value of 0.061. In contrast, 
the SPB method yields a p-value of 0.0192,
the NPB method yields a p-value of 0.0222, and the PB yields a p-value of
0.0141. Although such differences are not guaranteed, this
instance highlights a disagreement in conclusions among methods for the S\&P 500
index. Because NPB and PB both discard serial dependence, it is expected that
their conclusions could be different from our method. It is worth noting that
the SPB result is also not reliable here because the working dependence model is
in the class of ARMA dependence as specified in \citet{zeimbekakis2022misuses},
which does not cover nonlinear dependence usually captured by autoregressive
conditional heteroskedasticity \citep{engle1995arch}. The proposed method has
no need to specify the serial dependence and, hence, produces results under
weakest assumptions.

\section{Conclusion}
\label{sec:conclusion}

The KS test remains a crucial tool for researchers across diverse fields,
enabling critical assessments of population distributions. We introduced a
nonparametric block bootstrap variant of the KS test tailored for data with
serial dependencies, which eliminates the need for explicit modeling serial
dependence in addition to parameter specification within the hypothesized
distribution. Our simulations have demonstrated that, given
sufficiently large samples of time series data and moderate serial dependencies,
the test maintains its size under the null hypothesis and have considerable
power in detecting deviations from the hypothesized marginal distribution. The
method's performance improves as the strength of temporal dependence decreases
or the sample size increases. For moderate serial dependence levels in practice,
the test can be recommended for sample size~$n = 400$ or higher.
Overall, the test provides an analog of the bias-corrected nonparametric
bootstrap KS test of \citet{babu2004goodness} in the context of stationaly time
series with unspecified serial dependence. Compared to the semiparametric method
of \citet{zeimbekakis2022misuses}, which requires a working dependence model, it
is reliable regardless of whether the working dependence model approximates the
true dependence model.

Several promising directions for further research emerge from the findings of
this study. Firstly, analogous methodologies for other goodness-of-fit tests
based on empirical distributions, such as the Anderson--Darling and Cramér--von
Mises tests---which have been shown to outperform the KS test in certain scenarios
\citep{stephens2017tests}---could be explored. The application of block
bootstrapping bias correction to these tests could be particularly beneficial in
environmental studies, such as extreme value modeling, where observations are
often serially dependent. Secondly, an extension of the proposed nonparametric
block bootstrap one-sample KS test to a two-sample setting would be
intriguing. This adaptation would test whether two samples originate from the
same distribution without requiring specification of serial dependencies in
either sample, maintaining the advantageous feature of the original method.
